---
title: "impl_gavin"
output: html_document
---
## Multinomial Logistic Regression
This analysis implements multinomial logistic regression with elastic net regularization for predicting price tiers. The model uses tidymodels framework with parallel processing for hyperparameter tuning.

```{r load-libraries, include=FALSE}

library(tidyverse)
library(tidymodels)
library(patchwork)
library(nnet)
library(glmnet)
library(vip)
library(doParallel)
library(ggplot2)
library(gridExtra)
library(dplyr)

```


```{r setup-parallel, include=FALSE}

## Configure Parallel Processing
all_cores <- parallel::detectCores(logical = FALSE)
cl <- makePSOCKcluster(all_cores - 1)
registerDoParallel(cl)
```


```{r load-data-galvin, include=FALSE}
df_train <- read_csv("df_train.csv", show_col_types = FALSE)
df_test <- read_csv("df_test.csv", show_col_types = FALSE)

df_train$price_tier <- as.factor(df_train$price_tier)
df_test$price_tier <- as.factor(df_test$price_tier)

df_train <- df_train %>%
  mutate(
    across(where(is.logical), as.numeric),
    across(where(is.character), as.factor)
  )

df_test <- df_test %>%
  mutate(
    across(where(is.logical), as.numeric),
    across(where(is.character), as.factor)
  )

```

### Cross-Validation Setup and hyperparameter tuning
Here the model is tuned to find optimal hyperparameter using the 

```{r model-spec}

set.seed(123)
cv_folds <- vfold_cv(df_train, v = 5, strata = price_tier)

mlr_recipe <- recipe(price_tier ~ ., data = df_train) %>%
  step_rm(matches("^id$|^ID$"), skip = TRUE) %>%
  step_impute_median(all_numeric_predictors()) %>%
  step_impute_mode(all_nominal_predictors()) %>%
  step_nzv(all_predictors(), freq_cut = 95/5, unique_cut = 10) %>%
  step_normalize(all_numeric_predictors()) %>%
  step_corr(all_numeric_predictors(), threshold = 0.9) %>%
  step_other(all_nominal_predictors(), threshold = 0.01, other = "other") %>%
  step_integer(all_nominal_predictors()) %>%
  step_zv(all_predictors())

mlr_spec <- multinom_reg(
  penalty = tune(),
  mixture = tune()
) %>%
  set_engine("glmnet") %>%
  set_mode("classification")

mlr_workflow <- workflow() %>%
  add_recipe(mlr_recipe) %>%
  add_model(mlr_spec)


mlr_grid <- grid_regular(
  penalty(range = c(-4, 1)),
  mixture(range = c(0, 1)),
  levels = c(penalty = 10, mixture = 5)
)

# For multiclass classification, use yardstick metrics only
# Use only yardstick metrics explicitly to avoid caret conflicts
multiclass_metrics <- yardstick::metric_set(
  yardstick::accuracy,
  yardstick::roc_auc,
  yardstick::mn_log_loss,
  yardstick::sens,
  yardstick::spec,
  yardstick::precision,
  yardstick::f_meas)


tune_control <- control_grid(
  save_pred = TRUE,
  save_workflow = FALSE,
  verbose = TRUE,
  allow_par = TRUE,
  parallel_over = "everything",
  extract = NULL
)

start_time <- Sys.time()
set.seed(456)

mlr_tune_results <- tryCatch({
  tune_grid(
    mlr_workflow,
    resamples = cv_folds,
    grid = mlr_grid,
    metrics = multiclass_metrics,
    control = tune_control
  )
}, error = function(e) {
  cat("\nTuning error:", conditionMessage(e), "\n")
  return(NULL)
})

end_time <- Sys.time()
tuning_duration <- difftime(end_time, start_time, units = "mins")

if (is.null(mlr_tune_results)) {
  stopCluster(cl)
  registerDoSEQ()
  stop("Tuning failed. Check error messages.")
}
```

```{r tuning-results, include=FALSE}

all_metrics <- collect_metrics(mlr_tune_results)
cat("Total evaluations:", nrow(all_metrics), "\n")

if (nrow(all_metrics) == 0) {
  cat("\nNo successful models found.\n")
  print(collect_notes(mlr_tune_results))
  stopCluster(cl)
  registerDoSEQ()
  stop("All models failed.")
}
```

```{r viz-tuning-overview3, include=FALSE}
p1 <- autoplot(mlr_tune_results, metric = "accuracy") +
  labs(
    title = "Hyperparameter Tuning Results",
    subtitle = "5-Fold Cross-Validation",
    caption = paste("Combinations tested:", nrow(mlr_grid))
  ) +
  theme_minimal(base_size = 10) +  # Reduced base size for grid layout
  theme(
    plot.title = element_text(hjust = 0.5, size = 10),
    legend.title = element_blank(),
    legend.text = element_text(size = 8),
    axis.title = element_text(size = 9),
    axis.text = element_text(size = 8),
    axis.line = element_line(color = "black"),
    panel.grid = element_blank()
  )
```



### Performance by Regularization Parameters

```{r viz-performance-params, include=FALSE}
p2 <- mlr_tune_results %>%
  collect_metrics() %>%
  filter(.metric %in% c("accuracy", "roc_auc", "f_meas")) %>%
  ggplot(aes(x = penalty, y = mean, color = factor(mixture))) +
  facet_wrap(~.metric, scales = "free_y") +
  geom_line() +
  geom_point() +
  scale_x_log10() +
  labs(
    title = "Performance by Regularization Parameters",
    x = "Penalty (log scale)",
    y = "Mean Metric Value",
    color = "Mixture\n(0=Ridge, 1=Lasso)"
  ) +
  theme_minimal()
print(p2)
```


```{r viz-heatmap3, include=FALSE}
### Accuracy Heatmap
p3 <- mlr_tune_results %>%
  collect_metrics() %>%
  filter(.metric == "accuracy") %>%
  ggplot(aes(x = factor(round(mixture, 2)), 
             y = factor(round(log10(penalty), 2)), 
             fill = mean)) +
  geom_tile() +
  scale_fill_viridis_c() +
  labs(
    title = "Accuracy Heatmap",
    x = "Mixture (0=Ridge, 1=Lasso)",
    y = "log10(Penalty)",
    fill = "Mean\nAccuracy"
  ) +
  theme_minimal()
print(p3)
```

```{r best-model3}
best_params <- select_best(mlr_tune_results, metric = "roc_auc")
final_mlr_workflow <- finalize_workflow(mlr_workflow, best_params)
final_mlr_fit <- fit(final_mlr_workflow, data = df_train)

final_mlr_fit <- fit(final_mlr_workflow, data = df_train)
```


```{r train-performance, include=FALSE}

train_predictions <- predict(final_mlr_fit, df_train) %>%
  bind_cols(predict(final_mlr_fit, df_train, type = "prob")) %>%
  bind_cols(df_train %>% select(price_tier))

train_metrics <- train_predictions %>%
  metrics(truth = price_tier, estimate = .pred_class)


train_conf_mat <- train_predictions %>%
  conf_mat(truth = price_tier, estimate = .pred_class)

# autoplot(train_conf_mat, type = "heatmap") +
#   labs(title = "Training Set Confusion Matrix") +
#   theme_minimal()
```
### Test Set Performance

```{r test-performance}
# Get test predictions (class + probabilities)
test_predictions <- predict(final_mlr_fit, df_test) %>%
  bind_cols(predict(final_mlr_fit, df_test, type = "prob")) %>%
  bind_cols(df_test %>% select(price_tier))

# Compute metrics (using yardstick explicitly to avoid caret conflicts)
test_accuracy <- yardstick::accuracy(test_predictions, truth = price_tier, estimate = .pred_class)
test_sensitivity <- yardstick::sensitivity(test_predictions, truth = price_tier, estimate = .pred_class, estimator = "macro")
test_specificity <- yardstick::specificity(test_predictions, truth = price_tier, estimate = .pred_class, estimator = "macro")
test_precision <- yardstick::precision(test_predictions, truth = price_tier, estimate = .pred_class, estimator = "macro")
test_f1 <- yardstick::f_meas(test_predictions, truth = price_tier, estimate = .pred_class, estimator = "macro")

# Get probability columns for multiclass ROC AUC
prob_cols <- names(test_predictions)[stringr::str_detect(names(test_predictions), "^\\.pred_")]
prob_cols <- prob_cols[prob_cols != ".pred_class"]

test_roc_auc <- yardstick::roc_auc(test_predictions, truth = price_tier, all_of(prob_cols), estimator = "macro")

# Combine all metrics into a single summary table
test_metrics <- dplyr::bind_rows(
  test_accuracy,
  test_sensitivity,
  test_specificity,
  test_precision,
  test_f1,
  test_roc_auc
) %>%
  dplyr::select(.metric, .estimate) %>%
  dplyr::arrange(desc(.estimate))

# Display the summary
#print(test_metrics)

# Confusion matrix (heatmap visualization)
test_conf_mat <- yardstick::conf_mat(test_predictions, truth = price_tier, estimate = .pred_class)

p5 <- autoplot(test_conf_mat, type = "heatmap") +
  labs(
    title = "Test Set Confusion Matrix",
    subtitle = paste("Accuracy:", round(test_accuracy$.estimate, 4))
  ) +
  theme_minimal(base_size=10) +
  theme(
    plot.title = element_text(hjust = 0.5, size = 12),
    axis.title = element_text(size = 9),
    axis.text = element_text(size = 8),
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.title = element_text(size = 8),
    legend.text = element_text(size = 7)
  )

# Round estimates for readability
metrics_table <- test_metrics %>%
  arrange(desc(.estimate)) %>%
  mutate(.estimate = round(.estimate, 3))

# Create ggplot table
table_plot <- ggplot() +
  annotation_custom(
    grob = tableGrob(
      metrics_table,
      rows = NULL,          # removes row names
      theme = ttheme_default(
        core = list(fg_params = list(cex = 0.8)),   # text size
        colhead = list(fg_params = list(cex = 0.9, fontface = "bold"))
      )
    )
  ) +
  theme_void() +
  labs(title = "Classification Report") +
  theme(
    plot.title = element_text(
      size = 12, 
      hjust = 0.5,
      margin = margin(b = 5)
    ),
    plot.subtitle = element_text(
      size = 11,
      hjust = 0.5,
      margin = margin(b = 10)
    ),
    plot.margin = margin(10, 10, 10, 10)
  )

# Print the table
( p1 | p5 | table_plot) +
  plot_layout(ncol = 3, widths = c(1, 1, 1)) +
  plot_annotation(
    title = "MultiClass Logisitic Regression Model Performance Summary",
    theme = theme(
      plot.title = element_text(size = 12, face = "bold", hjust = 1)
    )
  )
```
Here, we have selected the optimal Regularized Multiclass Logistic Regression model by tuning the model using Nested Cross Validation:
 - The tuning architecture used had 5 outer folds to evaluate model performance and 5 inner folds to select optimal hyperparameters.
 - Tuned penalty (regularization strength) and alpha (regularizer type) using nested cross-validation.
 - Optimal hyperparameters found: penalty = 1e-4, alpha = 0.5 → elastic net regularization.
 
* Here, we see from the metrics table that even with the optimal hyperparameter and regularization, we see an accuracy of only 71% and a f1-score of 71% for the test data. This shows that the model does perform modestly but can be improved on even further.
* We also see that the model has a specificity of 85%, which shows that the model performs decently when identifying false classes for a test input, but falters to a 71% sensitivity when it has to correctly identify the correct class for the test data. This shows that the model is not a confident price tier identifier for new data, but it is a good false positive alarm.
```{r cleanup3, include=FALSE}
stopCluster(cl)
registerDoSEQ()
```