---
title: "Impl_Victor"
output: html_document
---

```{r load-libraries1, include=FALSE}
library(tidyverse)
library(patchwork)
library(grid)
library(gridExtra)
library(tidymodels)
library(ranger)
library(vip)
library(doParallel)
library(caret)

# Configure parallel processing
all_cores <- parallel::detectCores(logical = TRUE)
workers <- max(1, all_cores - 1)
cl <- makePSOCKcluster(workers)
registerDoParallel(cl)
```

```{r, include=FALSE}
# Load preprocessed data
df_train <- read_csv("df_train.csv", show_col_types = FALSE)
df_test <- read_csv("df_test.csv", show_col_types = FALSE)

# Ensure proper data types
df_train <- df_train %>%
  mutate(
    price_tier = factor(price_tier, levels = c("Budget", "Standard", "Premium")),
    across(where(is.character), as.factor)
  )

df_test <- df_test %>%
  mutate(
    price_tier = factor(price_tier, levels = c("Budget", "Standard", "Premium")),
    across(where(is.character), as.factor)
  )
```

## Random Forest model

This analysis implements a **Random Forest classifier** to predict Airbnb price tiers (Budget, Standard, Premium) in Melbourne. Random Forest is an ensemble method that builds multiple decision trees and aggregates their predictions, providing robust performance and natural resistance to overfitting.

**Target Distribution:**
```{r target-distribution, include=FALSE}
knitr::kable(
  prop.table(table(df_train$price_tier)) %>% 
    as.data.frame() %>%
    rename(Price_Tier = Var1, Proportion = Freq) %>%
    mutate(Proportion = round(Proportion, 3)),
  caption = "Training Set Class Distribution"
)
```


### Cross-Validation Strategy
- Used a **5-fold stratified cross-validation** to maintain class balance and random seed (123) for reproducibility
We also applied preprocessing steps for the workflow to be reproducible:
1. **Logical to numeric conversion**: TRUE/FALSE → 1/0 (required for ranger). 2. **Missing value imputation**: Median for numeric, mode for categorical. 3. **Zero-variance removal**: Eliminate uninformative constant features. 4. **Correlation filtering**: Remove highly correlated features (r > 0.9). 5. **Feature normalization**: Standardize numeric predictors for consistent importance metrics. 6. **Rare category grouping**: Combine categories with <1% frequency

### Model Specification

```{r model-spec-rf}
# Cross validation
set.seed(123)
cv_folds <- vfold_cv(df_train, v = 5, strata = price_tier)

# preprocess the data
rf_recipe <- recipe(price_tier ~ ., data = df_train) %>%
  step_rm(matches("^id$|^ID$"), skip = TRUE) %>%
  step_mutate(across(where(is.logical), as.numeric)) %>%
  step_impute_median(all_numeric_predictors()) %>%
  step_impute_mode(all_nominal_predictors()) %>%
  step_zv(all_predictors()) %>%
  step_corr(all_numeric_predictors(), threshold = 0.9) %>%
  step_normalize(all_numeric_predictors()) %>%
  step_other(all_nominal_predictors(), threshold = 0.01, other = "other_combined") %>% 
  step_dummy(all_nominal_predictors())

# define your model
rf_spec <- rand_forest(
  mtry = tune(),
  trees = tune(),
  min_n = tune()
) %>%
  set_engine(
    "ranger",
    importance = "impurity",
    seed = 123,
    num.threads = 1
  ) %>%
  set_mode("classification")

# Create a workflow
rf_workflow <- workflow() %>%
  add_recipe(rf_recipe) %>%
  add_model(rf_spec)

# define the metrics
multiclass_metrics <- yardstick::metric_set(
  yardstick::accuracy,
  yardstick::roc_auc,
  yardstick::mn_log_loss,
  yardstick::sens,
  yardstick::spec,
  yardstick::precision,
  yardstick::f_meas
)
```

**Hyperparameters tuned:**
- **mtry**: Number of features randomly sampled at each split (range: 10-30)
- **trees**: Number of trees in the forest (range: 500-2,000)
- **min_n**: Minimum observations in terminal nodes (range: 1-9). **Tuning strategy:** Latin hypercube sampling with 20 combinations for efficient parameter space coverage
```{r hyperparameter-tuning, echo=TRUE, message=TRUE, warning=FALSE}
# Check if tuning results already exist
if (file.exists("rf_tune_res.rds")) {
  #cat("✅ Loading saved tuning results...\n")
  rf_list <- readRDS("rf_tune_res.rds")
  rf_tune_results <- rf_list$tune_results
  rf_workflow <- rf_list$rf_workflow
} else {
  cat("⚙️ Running hyperparameter tuning (first time, may take a while)...\n")
  
  rf_grid <- grid_space_filling(
    parameters(
      mtry(range = c(5, 20)),
      trees(range = c(1000, 2000)),
      min_n(range = c(1, 9))
    ),
    size = 5,
    original = "lhs"
  )
  
  tune_control <- control_grid(
    save_pred = TRUE,
    save_workflow = FALSE,
    verbose = FALSE,
    allow_par = TRUE,
    parallel_over = "everything"
  )
  
  start_time <- Sys.time()
  set.seed(456)
  
  rf_tune_results <- tune_grid(
    rf_workflow,
    resamples = cv_folds,
    grid = rf_grid,
    metrics = multiclass_metrics,
    control = tune_control
  )
  
  rf_list <- list(
    tune_results = rf_tune_results,
    rf_workflow = rf_workflow
  )
  
  saveRDS(rf_list, file = "rf_tune_res.rds")
  end_time <- Sys.time()
  tuning_duration <- difftime(end_time, start_time, units = "mins")
  #cat("💾 Tuning complete and saved. Duration:", round(tuning_duration, 2), "mins\n")
}
```

```{r best-params, include=FALSE}
rf_list <- readRDS('rf_tune_res.rds')
rf_tune_results <- rf_list$tune_results
rf_workflow <- rf_list$rf_workflow
best_params <- select_best(rf_tune_results, metric = "roc_auc")

knitr::kable(
  best_params %>% 
    select(mtry, trees, min_n) %>%
    mutate(across(everything(), as.integer)),
  caption = "Best Hyperparameters (selected by ROC AUC)"
)
```

**Interpretation:**
- **mtry = `r best_params$mtry`**: Uses `r best_params$mtry` features per split (out of ~`r ncol(df_train)-1` total)
- **trees = `r best_params$trees`**: Ensemble of `r best_params$trees` decision trees
- **min_n = `r best_params$min_n`**: Minimum `r best_params$min_n` observations required to split a node

```{r cv-performance, include=FALSE}
cv_summary <- tibble(
  Metric = c("ROC AUC", "Accuracy", "F1 Score"),
  CV_Mean = c(
    show_best(rf_tune_results, metric = "roc_auc", n = 1)$mean,
    show_best(rf_tune_results, metric = "accuracy", n = 1)$mean,
    show_best(rf_tune_results, metric = "f_meas", n = 1)$mean
  )
) %>%
  mutate(
    CV_Mean = round(CV_Mean, 4)
  )
```

### Hyperparameter Tuning Visualization

```{r fig-tuning-results, include=FALSE, fig.cap="Model performance across different hyperparameter combinations. Each point represents a unique configuration."}
autoplot(rf_tune_results, metric = "roc_auc") +
  labs(
    title = "Random Forest Hyperparameter Tuning Results",
    subtitle = "5-Fold Cross-Validation - ROC AUC"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5)
  )

rf_tune_results %>%
  collect_metrics() %>%
  filter(.metric %in% c("accuracy", "roc_auc")) %>%
  ggplot(aes(x = trees, y = mean, color = factor(mtry))) +
  facet_wrap(~.metric, scales = "free_y") +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_color_viridis_d(option = "plasma", end = 0.9) +
  labs(
    title = "Impact of Number of Trees on Performance",
    x = "Number of Trees",
    y = "Mean Metric Value",
    color = "mtry"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    strip.text = element_text(face = "bold"),
    plot.title = element_text(hjust = 0.5, face = "bold"),
    legend.position = "bottom"
  )

g1 <- rf_tune_results %>%
  collect_metrics() %>%
  filter(.metric == "accuracy") %>%
  ggplot(aes(x = mtry, y = mean)) +
  geom_point(aes(size = trees, color = min_n), alpha = 0.7) +
  geom_smooth(method = "loess", se = TRUE, color = "darkblue", linewidth = 1) +
  scale_color_viridis_c(option = "mako") +
  scale_size_continuous(range = c(2, 8)) +
  labs(
    title = "Feature Sampling (mtry) Impact on Accuracy",
    x = "Number of Features per Split (mtry)",
    y = "Mean Accuracy",
    color = "min_n",
    size = "trees"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold")
  )
```

**Key Findings:**
- Model performance `r ifelse(max(cv_summary$CV_Mean) - min(cv_summary$CV_Mean) < 0.05, "is relatively stable", "varies significantly")` across different hyperparameter combinations
- `r ifelse(best_params$trees > 1500, "High", ifelse(best_params$trees < 750, "Low", "Moderate"))` number of trees (`r best_params$trees`) suggests `r ifelse(best_params$trees > 1500, "ensemble diversity is beneficial", "efficiency-accuracy trade-off")`
- `r ifelse(best_params$mtry < 15, "Low", ifelse(best_params$mtry > 25, "High", "Moderate"))` mtry value indicates `r ifelse(best_params$mtry < 15, "high randomness per tree", ifelse(best_params$mtry > 25, "using most features per split", "balanced feature sampling"))`
```{r train-final}
final_rf_workflow <- finalize_workflow(rf_workflow, best_params)
final_rf_fit <- fit(final_rf_workflow, data = df_train)
final_rf_model <- extract_fit_parsnip(final_rf_fit)

test_predictions <- predict(final_rf_fit, df_test) %>%
  bind_cols(predict(final_rf_fit, df_test, type = "prob")) %>%
  bind_cols(df_test %>% select(price_tier))

# Calculate all metrics
test_accuracy <- yardstick::accuracy(test_predictions, truth = price_tier, estimate = .pred_class)
test_sensitivity <- yardstick::sensitivity(test_predictions, truth = price_tier, estimate = .pred_class, estimator = "macro")
test_specificity <- yardstick::specificity(test_predictions, truth = price_tier, estimate = .pred_class, estimator = "macro")
test_precision <- yardstick::precision(test_predictions, truth = price_tier, estimate = .pred_class, estimator = "macro")
test_f1 <- yardstick::f_meas(test_predictions, truth = price_tier, estimate = .pred_class, estimator = "macro")

prob_cols <- names(test_predictions)[stringr::str_detect(names(test_predictions), "^\\.pred_")]
prob_cols <- prob_cols[prob_cols != ".pred_class"]
test_roc_auc <- yardstick::roc_auc(test_predictions, truth = price_tier, all_of(prob_cols), estimator = "hand_till")

test_logloss <- yardstick::mn_log_loss(test_predictions, truth = price_tier, all_of(prob_cols))


```


```{r test-metrics-table, include=FALSE}
test_metrics <- bind_rows(
  test_accuracy,
  test_roc_auc,
  test_logloss,
  test_sensitivity,
  test_specificity,
  test_precision,
  test_f1
) %>%
  select(.metric, .estimate) %>%
  mutate(
    .metric = recode(.metric,
      "accuracy" = "Accuracy",
      "roc_auc" = "ROC AUC",
      "mn_log_loss" = "Log Loss",
      "sens" = "Sensitivity (Macro)",
      "spec" = "Specificity (Macro)",
      "precision" = "Precision (Macro)",
      "f_meas" = "F1 Score (Macro)"
    ),
    .estimate = round(.estimate, 4)
  ) %>%
  rename(Metric = .metric, Value = .estimate)

knitr::kable(
  test_metrics,
  caption = "Final Model Performance on Test Set (n = 4,580)",
  align = c('l', 'r')
)
```

```{r caret-confusion, include=FALSE}
# Generate caret confusion matrix for detailed metrics
caret_cm <- caret::confusionMatrix(
  test_predictions$.pred_class,
  df_test$price_tier
)

# Extract per-class metrics
class_stats <- as.data.frame(caret_cm$byClass) %>%
  rownames_to_column("Class") %>%
  mutate(Class = str_remove(Class, "Class: ")) %>%
  select(
    Class,
    Sensitivity,
    Specificity,
    Precision = `Pos Pred Value`,
    F1,
    `Balanced Accuracy`
  ) %>%
  mutate(across(where(is.numeric), ~round(., 3)))

knitr::kable(
  class_stats,
  caption = "Per-Class Performance Metrics"
)

test_conf_mat <- yardstick::conf_mat(test_predictions, truth = price_tier, estimate = .pred_class)

# Display as table
conf_mat_table <- test_conf_mat$table %>%
  as.data.frame() %>%
  pivot_wider(names_from = Prediction, values_from = Freq)

knitr::kable(
  conf_mat_table,
  caption = "Confusion Matrix: True Labels (rows) vs Predicted Labels (columns)"
)

```

```{r fig-confusion-heatmap,include=FALSE, fig.cap="Confusion matrix heatmap. Darker colors on the diagonal indicate correct predictions."}
autoplot(test_conf_mat, type = "heatmap") +
  labs(
    title = "Test Set Confusion Matrix",
    subtitle = paste0("Overall Accuracy: ", round(test_accuracy$.estimate, 3))
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5)
  )
```


```{r confusion-insights, results='asis', include=FALSE}
# Calculate misclassification patterns
conf_data <- test_conf_mat$table %>% as.data.frame()

# Find most common misclassification
misclass <- conf_data %>%
  filter(Truth != Prediction) %>%
  arrange(desc(Freq)) %>%
  head(1)

cat(paste0(
  "- Most common error: Predicting **", misclass$Prediction, 
  "** when true class is **", misclass$Truth, 
  "** (", misclass$Freq, " cases)\n"
))

# Calculate diagonal accuracy
diag_sum <- sum(diag(test_conf_mat$table))
total <- sum(test_conf_mat$table)
cat(paste0("- Correctly classified: ", diag_sum, " out of ", total, " observations (", 
           round(diag_sum/total*100, 1), "%)\n"))

# Class-specific insights
for(class in levels(df_test$price_tier)) {
  class_total <- sum(test_conf_mat$table[class, ])
  class_correct <- test_conf_mat$table[class, class]
  class_acc <- round(class_correct / class_total, 3)
  cat(paste0("- **", class, "** tier: ", class_acc, " accuracy (", 
             class_correct, "/", class_total, ")\n"))
}
```

### ROC Curves and parameter tuning visualization

```{r fig-roc-curves, fig.cap="ROC curves for each price tier using one-vs-all strategy. Curves closer to top-left indicate better discrimination.", warning=FALSE}

roc_data <- test_predictions %>%
  yardstick::roc_curve(truth = price_tier, all_of(prob_cols))

# autoplot(roc_data) +
#   labs(
#     title = "ROC Curves by Price Tier (One-vs-All)",
#     subtitle = paste0("Overall AUC (Hand & Till): ", round(test_roc_auc$.estimate, 4))
#   ) +
#   theme_minimal(base_size = 12) +
#   theme(
#     plot.title = element_text(hjust = 0.5, face = "bold"),
#     plot.subtitle = element_text(hjust = 0.5),
#     legend.position = "bottom"
#   )

g3 <- roc_data %>%
  ggplot(aes(x = 1 - specificity, y = sensitivity, color = .level)) +
  geom_path(linewidth = 1) +
  geom_abline(lty = 2, color = "gray50") +
  coord_equal() +
  scale_color_viridis_d(option = "plasma", end = 0.9) +
  labs(
    title = "ROC Curves by Price Tier (One-vs-All)",
    subtitle = paste0("Overall AUC (Hand & Till): ", round(test_roc_auc$.estimate, 4)),
    x = "False Positive Rate (1 - Specificity)",
    y = "True Positive Rate (Sensitivity)",
    color = "Price Tier"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5),
    legend.position = "bottom"
  )

g3+g1

```

```{r generalization-check}
comparison <- tibble(
  Metric = c("ROC AUC", "Accuracy", "F1 Score"),
  `Cross-Validation` = c(
    show_best(rf_tune_results, metric = "roc_auc", n = 1)$mean,
    show_best(rf_tune_results, metric = "accuracy", n = 1)$mean,
    show_best(rf_tune_results, metric = "f_meas", n = 1)$mean
  ),
  `Test Set` = c(
    test_roc_auc$.estimate,
    test_accuracy$.estimate,
    test_f1$.estimate
  )
) %>%
  mutate(
    Difference = `Test Set` - `Cross-Validation`,
    `% Change` = round((Difference / `Cross-Validation`) * 100, 2),
    across(c(`Cross-Validation`, `Test Set`), ~round(., 4))
  )

# knitr::kable(
#  comparison,
#  caption = "Cross-Validation vs Test Set Performance Comparison")

g4 <- gridExtra::tableGrob(
  comparison,
  rows = NULL,
  theme = gridExtra::ttheme_minimal(
    core = list(fg_params = list(hjust = 0, x = 0.1)),
    colhead = list(fg_params = list(fontface = "bold"))
  )
)

# Add title to the table
g4 <- gridExtra::arrangeGrob(
  g4,
  top = grid::textGrob(
    "Cross-Validation vs Test Set Performance",
    gp = grid::gpar(fontsize = 12, fontface = "bold")
  )
)

```

```{r feature-importance, include=FALSE}
importance_scores <- final_rf_model %>%
  extract_fit_engine() %>%
  vip::vi()
```

#### Most Important Feature analysis and model generalization

```{r importance-table, include=FALSE}
knitr::kable(
  head(importance_scores, 20) %>%
    mutate(
      Importance = round(Importance, 2),
      Rank = row_number()
    ) %>%
    select(Rank, Variable, Importance),
  caption = "Top 20 Features by Importance (Gini Impurity)"
)
```

```{r fig-feature-importance, fig.cap="Variable importance based on mean decrease in Gini impurity. Higher values indicate more influential features."}
g2 <- importance_scores %>%
  slice_max(order_by = Importance, n = 20) %>%
  ggplot(aes(x = Importance, y = reorder(Variable, Importance))) +
  geom_col(fill = "steelblue", alpha = 0.8) +
  geom_text(aes(label = round(Importance, 1)), hjust = -0.1, size = 3) +
  labs(
    title = "Top 20 Most Important Features",
    subtitle = "Random Forest - Gini Impurity Importance",
    x = "Importance Score",
    y = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5),
    panel.grid.major.y = element_blank()
  ) +
  xlim(0, max(importance_scores$Importance[1:20]) * 1.1)


g2+wrap_elements(g4)
```
**Generalization Assessment:**
`r if(abs(comparison$"% Change"[1]) < 2) {"✓ **Excellent generalization** - CV and test performance are nearly identical (<2% difference)"} else if(abs(comparison$"% Change"[1]) < 5) {"✓ **Good generalization** - Minor difference between CV and test (<5%)"} else {"⚠️ **Potential overfitting** - Notable gap between CV and test performance (>5%)"}`. The Random Forest model `r if(comparison$"% Change"[1] > 0) {"performs slightly better"} else {"performs slightly worse"}` on the test set compared to cross-validation, with a `r abs(comparison$"% Change"[1])`% difference in ROC AUC.

### Model Performance Summary

The Random Forest classifier achieved:
- **`r round(test_accuracy$.estimate * 100, 1)`% overall accuracy** on the test set
- **`r round(test_roc_auc$.estimate, 3)` ROC AUC** indicating `r ifelse(test_roc_auc$.estimate > 0.85, "excellent", ifelse(test_roc_auc$.estimate > 0.75, "good", "moderate"))` discrimination ability
- **`r round(test_f1$.estimate, 3)` F1 score** demonstrating balanced precision and recall
- **`r abs(round(comparison$"% Change"[1], 1))`% difference** between CV and test performance (strong generalization)

```{r cleanup, include=FALSE}
stopCluster(cl)
registerDoSEQ()
```

